### 2012-04-26 3.1.1.tex
Albright MS data and some writing on chapter 3

One of the most salient properties of Turkish is its harmony system, which \citeauthor{Lees1966b} (\citeyear[][35]{Lees1966b}, \citeyear[][284]{Lees1966a}) derives by means of feature spreading rules. The least constrained of these processes is the rule of backness harmony described in this section.

@@

Backness harmony spreads the value of \textsc{Back} rightward over any intervening consonants onto the next vowel.

@@

\xymatrix@R=24pt@C=24pt{
\txt{V}                                        & \txt{C$_0$} & \txt{V} \\
\txt{[α \textsc{Back}]}\ar@{-}[u]\ar@{..}[urr] \\
}
\end{example}

Along with the rule of \textsc{Roundness Harmony} discussed in the next section, this rule accounts for a great deal of the suffix allomorphy found in Turkish. The nominative plural (nom.pl.) is formed by adding the suffixes \emph{-ler} or \emph{-lar} to the nominative singular (nom.sg.).

\begin{example}
The Turkish nominative:

@@

\noindent
\textsc{Backness Harmony} correctly predicts the choice of suffix. There are a few complications, however. First, while most polysyllabic roots conform to \textsc{Backness Harmony} (\lastx b), not all do. In this case, suffix vowels generally harmonize with the final root vowel (\nextx a). Yet there is also a small class of nouns which exhibit [$-$\textsc{Back}] suffix vowels despite the fact that their final root vowel is [$+$\textsc{Back}] (\nextx b).

\begin{example}
Exceptional Turkish nominatives:

@@

While it is uncontroversial that disharmonic suffixes (\lastx b) are no more than sporadic exceptions to \textsc{Backness Harmony}, the status of root disharmony (\lastx a) has been the subject of much debate. \citet[][212, 289]{Anderson1974} views the fact that disharmonic roots still trigger suffix harmony as evidence that suffix harmony is distinct from root harmony, the latter being a sequence structure constraint.

@@

Underspecification provides an alternative under which the exceptionality of disharmonic roots is only apparent. While it is clear that the first vowel of Turkish roots are contrastively specified for backness (e.g., \emph{kül} `ash' vs.  \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'), there are also some minimal pairs with regards to backness (dis)harmony (e.g., \emph{deve} `camel' vs. \emph{deva} `medicine', \emph{sene} `year' vs. \emph{sena} `praise'). This supports the possibility, first suggested by \citet{Clements1982}, that harmonic roots may be underspecified and disharmonic roots fully specified, exemplified below.

@@

\begin{example}
Underlying specification of \textsc{Back} for (dis)harmonic roots:

\xymatrix@R=24pt@C=24pt{
\txt{a.} & \txt{s} & \txt{V} & \txt{n} & \txt{V}  & \txt{\emph{sene} `year'} \\
&        & \txt{[$-$\textsc{Back}]}\ar@{-}[u]     \\
\txt{b.} & \txt{s} & \txt{V} & \txt{n} & \txt{V} & \txt{\emph{sena} `praise'} \\
         &         & \txt{[$-$\textsc{Back}]}\ar@{-}[u] & & \txt{[$+$\textsc{Back}]}\ar@{-}[u]
}
\end{example}

@@

However, the theory of exceptionality presented in \emph{SPE} (p.~374f.) provides a direct account of suffix harmony in disharmonic roots without duplication. In \emph{SPE}, the specification of the target (i.e., the segment to be changed) of every rule $R$ must be [$+$rule $R$] by convention.  A root or affix which fails to undergo $R$ despite otherwise matching the structural description is simply said to be marked [$-$rule $R$], and thus failing to meet the full structural description. If disharmonic roots are [$-$\textsc{Backness Harmony}], then suffix vowels are correctly predicted to undergo \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the environment, which is not required to be [$+$rule $R$].

@@

In both Tuvan and Turkish, the unchanged reduplicant vowel is also affected. Reduplication interacts with root harmony in both Tuvan and Turkish. When the base is harmonically [$-$\textsc{Back}], the insertion of a [$+$\textsc{Back}] results in what \citeauthor{Harrison2001} call ``reharmonization'' (\nextx a).

@@

\noindent
Under the underspecification hypothesis, non-initial harmonic vowels lack an underlying \textsc{Back} feature, so it comes as no surprise that changing the \textsc{Back} specification of the root-initial vowel results in reharmonization. And the full specification of disharmonic roots correctly predicts that they will be exempt from reharmonization (\lastx b), which is also borne out in Tuvan and in an unrelated Finnish language game \citep{Campbell1986}.

>>>>

One of the most salient properties of Turkish is its harmony system, which \citeauthor{Lees1966b} (\citeyear[][35]{Lees1966b}, \citeyear[][284]{Lees1966a}) derives by means of feature spreading rules. Backness harmony is the the most general of these processes.

@@

The rule of \textsc{Backness Harmony} spreads a vowel's \textsc{Back} specification rightward over any intervening consonants onto the next vowel.

@@

\xymatrix@R=24pt@C=24pt{
\txt{V}                                        & \txt{C$_0$} & \txt{V} \\
\txt{[α \textsc{Back}]}\ar@{-}[u]\ar@{--}[urr] \\
}
\end{example}

This rule accounts for a great deal of the suffix allomorphy found in Turkish. For instance, the nominative plural (nom.pl.) suffix is \emph{-ler} when the final root vowel is [$-$\textsc{Back}], and \emph{-lar} when it is [$+$\textsc{Back}], as predicted by Backness Harmony.

\begin{example}
\label{turknom}
The Turkish nominative:

@@

There are however a few complications. First, not all polysyllabic roots conform to \textsc{Backness Harmony}. In this case, as can be seen from (\ref{turkexcept}a), suffix vowels generally harmonize with the final root vowel. 
There is also a small class of nouns, shown in (\ref{turkexcept}b), which exhibit [$-$\textsc{Back}] suffix vowels despite the fact that their final root vowel is [$+$\textsc{Back}].

\begin{example}
\label{turkexcept}
Exceptional Turkish nominatives:

@@

It is uncontroversial that the disharmonic suffixes of (\ref{turkexcept}b) are no more than very sporadic exceptions to \textsc{Backness Harmony}, root disharmony has ben the subject of much debate.  As disharmonic roots still trigger suffix harmony, \citet[][212, 289]{Anderson1974} proposes to distinguish suffix harmony (an alternation) from a sequence structure constraint governing root harmony.

However, as \citet[][?]{Zonneveld1978} notes, the theory of exceptionality proposed in \emph{SPE} (p.~374f.) provides a direct account of suffix harmony in disharmonic roots. \citeauthor{SPE} assume that the specification of the target (i.e., the segment or segments to be changed) of a rule $R$ must be marked [$+$rule $R$] by convention. A root or affix which fails to undergo $R$ despite otherwise matching the structural description is simply said to be marked [$-$rule $R$]. Under this account, exceptionality boils down to failing to match an extended structural description. If disharmonic roots are marked [$-$\textsc{Backness Harmony}], then the final vowel of disharmonic roots will still trigger \textsc{Backness Harmony}, since the [$-$\textsc{Backness Harmony}] root is no longer the target but rather the environment, which is not subject to the [$+$\textsc{Backness Harmony}] requirement.

@@

An alternative account, first proposed by \citet{Clements1982}, uses underspecificatio to derive the non-exceptionality of harmonic roots. The initial-syllable vowel of a Turkish root is contrastively specified for backness (e.g., \emph{kül} `ash' vs.  \emph{kul} `servant', \emph{kepek} `bran' vs. \emph{kapak} `lid'); harmonic roots are ones in which non-initial syllable vowels cannot contrast for this feature, and \citeauthor{Clements1982} proposes to leave them underspecified. Only in disharmonic roots does \textsc{Back} need to be specified for non-initial syllable vowels. 
%This is exemplified below in (\ref{spec}).
%(e.g., \emph{deve} `camel' vs. \emph{deva} `medicine', \emph{sene} `year' vs. \emph{sena} `praise').

@@

\begin{example}
\label{spec}
\textsc{Back} specification for (dis)harmonic roots:

\vspace{0.5\baselineskip}
\xymatrix@R=24pt@C=24pt{
\txt{a.} & \txt{harmonic root:~~~~} & \txt{C} & \txt{V} & \txt{C} & \txt{V} \\
&   &    & \txt{[$-$\textsc{Back}]}\ar@{-}[u]\ar@{--}[urr] \\
\txt{b.} & \txt{disharmonic root:} & \txt{C} & \txt{V} & \txt{C} & \txt{V} \\
    &    &         & \txt{[$-$\textsc{Back}]}\ar@{-}[u] & & \txt{[$+$\textsc{Back}]}\ar@{-}[u]
}
\end{example}

@@

\citet{Clements1982} and \citet{Kaun1999}

\footnote{Thanks to Kie Zuraw for bringing this data to my attention.}

the epenthetic vowel is always [$+$\textsc{High}], and when the first consonant of the cluster is non-dorsal, it shows backness harmony

\begin{example}
Adaptation of initial foreign clusters \citep[][247]{Clements1982}:

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l l}
a. & spiker  & \alt{} & sipiker  & `announcer' \\
   & fren    & \alt{} & firen    & `break'     \\
   & trablus & \alt{} & tırablus & `Tripoli'   \\
   & brom    & \alt{} & burom    & `bromide'   \\
   & prusya  & \alt{} & purusya  & `Prussia'   \\
b. & kral    & \alt{} & kıral    & `king'      \\
   & grup    & \alt{} & gurup    & `group'     \\
c. & grip    & \alt{} & gırip    & `grippe'    \\ % unexpectedly back
   & kredi   & \alt{} & kıredi   & `credit'    \\
\end{tabular}
\end{example}

@@

In both Tuvan and Turkish, the unchanged reduplicant vowel is also affected. Reduplication interacts with root harmony in both Tuvan and Turkish. When the base is harmonically [$-$\textsc{Back}], the insertion of a [$+$\textsc{Back}] results in what \citeauthor{Harrison2001} call ``reharmonization'' (\ref{redupgame}a).

@@

\noindent
Under the underspecification hypothesis, non-initial harmonic vowels lack an underlying \textsc{Back} feature, so it comes as no surprise that changing the \textsc{Back} specification of the root-initial vowel results in reharmonization. And the full specification of disharmonic roots correctly predicts that they will be exempt from reharmonization (\ref{redupgame}b), which is also borne out in Tuvan and in an unrelated Finnish language game \citep{Campbell1986}.

### 2012-04-26 3.1.tex
Albright MS data and some writing on chapter 3

\citet{Lees1966a,Lees1966b} proposes three sequence structure constraints on Turkish vowels, and these constraints have been discussed extensively in the subsequent literature. Below, these constraints are formalized and related to Turkish alternation phonology and various sources of external evidence. A traditional binary feature specification for the Turkish vowel system, given below, is assumed without discussion.

@@

\vspace{0.5\baselineskip}
\begin{tabular}{c | c c c c}
                       & \multicolumn{2}{c}{[$-$\textsc{Back}]} & \multicolumn{2}{c}{[$+$\textsc{Back}]} \\
                       & [$-$\textsc{Round}] & [$+$\textsc{Round}] & [$-$\textsc{Round}] & [$+$\textsc{Round}] \\ \midrule
\buf[$+$\textsc{High}] & i & ü [y] & ı [ɯ] & u \\
\buf[$-$\textsc{High}] & e & ö [ø] & a [ɑ] & o \\
\end{tabular}
\end{example}

>>>>

\citet{Lees1966a,Lees1966b} proposes three constraints on Turkish vowel sequences 
which have been the attention of many subsequent studies on vowel harmony and on sequence structure constraints. In this section, these constraints are formalized, and where possible, related to phonological alternations in Turkish and to various sources of external evidence. The following feature specification for the Turkish vowels is assumed throughout.

@@

\vspace{0.5\baselineskip}
\begin{tabular}{c | c c c c}
                       & \multicolumn{2}{c}{[$-$\textsc{Back}]} & \multicolumn{2}{c}{[$+$\textsc{Back}]} \\
                       & [$-$\textsc{Round}] & [$+$\textsc{Round}] & [$-$\textsc{Round}] & [$+$\textsc{Round}] \\ \midrule
\buf[$+$\textsc{High}] & {i} & {ü} [y] & {ı} [ɯ] & {u} \\
\buf[$-$\textsc{High}] & {e} & {ö} [ø] & {a} [ɑ] & {o} \\
\end{tabular}
\end{example}

### 2012-04-26 3.tex
Albright MS data and some writing on chapter 3

\chapter{Static and derived constraints in Turkish}
\label{turkish}

@@

The contrast between \emph{blick} and \emph{bnick} is what has been called a "projection problem" \citep{Baker1979}

This is not a universal, but rather something English speakers must learn, as some languages, such as Moroccan Arabic, do in fact permit \emph{bn} onsets (e.g., \emph{bni} 'building', \emph{bnat} 'daughters', \emph{bniqa} 'closet');

@@

% Berent et al. 2007
% 598: Highly marked inputs are repaired in perception to abide by the grammatical restrictions of the language
% Exps 1-2: syllable count task (misperception due to linguistic experience; "disyllables were more likely to be perceived as monosyllables if their monosyllabic counterparts had a less marked cluster" in both Russian and English, not what OT predicts)
% Exps 3-4: AB task
% Exps 5-6: priming effect (non-epenthetic and epenethetic things represented the same)
% Berent and Lennertz 2007: (640: If the high rate of monosyllabic errors to unmarked onsets is only due to phonetic failures to encode the input, then it is puzzling why the same onsets also yield the highest rate of accurate response.)
% Berent et al. 2008a:
\citet{Peperkamp2007}
% Peperkamp 2007: during one stage of processing phonology is undone with help from the lexicon...it only takes licit inputs and is irrelevant (p. 633)
\citet{Dupoux1999}
% Dupoux et al. 1999: Japanese speakers still can't distinguish between ebzo and ebuzo in an ABX task using the SAME a or b.

>>>>

\chapter{Static and derived constraints}
\label{turkish}

@@

The contrast between \emph{blick} and \emph{bnick} is what has been called a ``projection problem'' \citep{Baker1979}

This is not a universal, but rather something English speakers must learn: Moroccan Arabic, for instance, does permit \emph{bn} onsets (e.g., \emph{bni} 'building', \emph{bnat} 'daughters', \emph{bniqa} 'closet').

@@

% Berent et al. 2007
% 598: Highly marked inputs are repaired in perception to abide by the grammatical restrictions of the language
% Exps 1-2: syllable count task (misperception due to linguistic experience; "disyllables were more likely to be perceived as monosyllables if their monosyllabic counterparts had a less marked cluster" in both Russian and English, not what OT predicts)
% Exps 3-4: AB task
% Exps 5-6: priming effect (non-epenthetic and epenethetic things represented the same)
% Berent and Lennertz 2007: (640: If the high rate of monosyllabic errors to unmarked onsets is only due to phonetic failures to encode the input, then it is puzzling why the same onsets also yield the highest rate of accurate response.)
% Berent et al. 2008a:
\citet{Peperkamp2007}
% Peperkamp 2007: during one stage of processing phonology is undone with help from the lexicon...it only takes licit inputs and is irrelevant (p. 633)
\citet{Dupoux1999}
% Dupoux et al. 1999: Japanese speakers still can't distinguish between ebzo and ebuzo in an ABX task using the SAME a or b.
%Elliot Moreton (p.c.)

### 2012-04-26 4.1.2.tex
Albright MS data and some writing on chapter 3

It is necessary to split medial consonant clusters into the codas and onsets that make them up. While CELEX provides syllabified transcriptions, no description of the syllabification procedure is given in the documentation, and inspection of these syllabifications suggests the procedure used is not fully systematic. For instance, \emph{chemistry} is syllabified as [ˈkɛ.mɪ.strɪ] but \emph{ministry} as [ˈmɪ.nɪs.trɪ].\footnote{Note that word-final \emph{y} is generally lax [ɪ] in Received Pronunciation \citep[][294]{AOE2}.} This putative [ɪ.strɪ $\sim$ ɪs.trɪ] contrast, and many other such contrasts in the CELEX data, are dubious simply because there is no clear evidence that contrastive syllabification occurs in any language. Apparent counterexamples appear to be effects of vowel or consonant length \citep[e.g.,][]{Elfner2006}, word stress, or morphological structure, none of which contrast \emph{chemistry} and \emph{ministry}.

@@

There is one case where unchecked maximization of medial onsets produces incorrect syllabifications. When a medial consonant cluster is preceded by a stressed lax vowel, as in words like \emph{propaga}[n.d]\emph{a}, \emph{whi}[s.p]\emph{er}, \emph{vi}[s.t]\emph{a}, or \emph{bi}[s.k]\emph{uit}, the first consonant of the cluster checks the lax vowel \citep[e.g.,][3]{Hammond1997}. Onset maximization makes the right predictions for words like \emph{propaganda}, where the medial cluster cannot be an onset, but \citet[][55]{Harris1994} notes that it makes the wrong predictions when the medial cluster is a valid onset: in \emph{whisper}, \emph{vista}, and \emph{biscuit}, it incorrectly assigns [sp, st, sk] to the medial onset, leaving the coda empty. The approach adopted here is to first apply onset maximization, then to reassign the first consonant of a medial onset to the preceding coda if it is immediately preceded by a stressed lax vowel.

>>>>

For this task, it is necessary to separate medial consonant clusters into codas and onsets. While CELEX provides syllabified transcriptions, no description of the syllabification procedure is given in the documentation, and inspection of these syllabifications suggests the procedure used is not fully systematic. For instance, \emph{chemistry} is syllabified as [ˈkɛ.mɪ.strɪ] but \emph{ministry} as [ˈmɪ.nɪs.trɪ].\footnote{Note that word-final \emph{y} is generally lax [ɪ] in Received Pronunciation \citep[][294]{AOE2}.} This putative [ɪ.strɪ $\sim$ ɪs.trɪ] contrast, and many other such contrasts in the CELEX data, are dubious simply because there is no clear evidence that contrastive syllabification occurs in any language. Apparent counterexamples appear to be effects of vowel or consonant length \citep[e.g.,][]{Elfner2006}, word stress, or morphological structure, none of which explain the \emph{chemistry}/\emph{ministry} syllabification contrast.

@@

There is one case where unchecked maximization of medial onsets produces incorrect syllabifications. When a medial consonant cluster is preceded by a stressed lax vowel, as in words like \emph{propaga}[n.d]\emph{a}, \emph{whi}[s.p]\emph{er}, \emph{vi}[s.t]\emph{a}, or \emph{bi}[s.k]\emph{uit}, the first consonant of the cluster checks the lax vowel \citep[e.g.,][3]{Hammond1997}. \citet[][55]{Harris1994} notes that onset maximization makes the wrong predictions when the medial cluster is a valid onset: in \emph{whisper}, \emph{vista}, and \emph{biscuit}, it incorrectly assigns [sp, st, sk] to the medial onset, leaving the coda empty. The approach adopted here is to first apply onset maximization, then to reassign the first consonant of a medial onset to the preceding coda if it is immediately preceded by a stressed lax vowel.

### 2012-04-26 4.1.3.tex
Albright MS data and some writing on chapter 3

\begin{example}
VBR's complex onsets \citep[79--80, his transcriptions]{Buchwald2005}:

@@

\noindent
VBR breaks up complex onsets with epenthesis, including those that consist of a consonant and a back onglide cluster (\lastx b). However, no epenthesis occurs between a consonant and a front onglide; rather, the glide is absent (\lastx c). The failure of the front onglide to pattern with other consonant clusters suggests once again that the glide is part of the nucleus.

@@

\noindent
Similarly, while /t, d/ delete in word-final position when immediately preceded by a consonant, including sonorants /n, l/, as in (\nextx a), deletion of /t, d/ after post-vocalic \emph{r} in American English is ``rare or nonexistent'' \citep[][8]{Guy1980}.

\begin{example}
/t, d/-\textsc{Deletion} in American English:

>>>>

\begin{example}
\label{VBR}
VBR's complex onsets \citep[79--80, his transcriptions]{Buchwald2005}:

@@

\noindent
VBR breaks up complex onsets with epenthesis, including those that consist of a consonant and a back onglide cluster (\ref{VBR}b). However, no epenthesis occurs between a consonant and a front onglide; rather, the glide is absent (\ref{VBR}c). The failure of the front onglide to pattern with other consonant clusters suggests once again that the glide is part of the nucleus.

@@

\noindent
Similarly, while /t, d/ delete in word-final position when immediately preceded by a consonant, including sonorants /n, l/, as in (\ref{td}a), deletion of /t, d/ after post-vocalic \emph{r} in American English is ``rare or nonexistent'' \citep[][8]{Guy1980}.

\begin{example}
\label{td}
/t, d/-\textsc{Deletion} in American English:

### 2012-04-26 4.2.1.tex
Albright MS data and some writing on chapter 3

Above, it was assumed that the velar nasal is derived from underlying /n/. 
This analysis is not purely allophonic, as nasal place assimilation alternations target the final consonant of the \emph{im-}/\emph{in-} prefix. In (\lastx ab), the prefix-final consonant takes on the same major place of articulation as the following obstruent, and in (\lastx c), where the root is vowel-initial, the prefix-final consonant is realized as a default /n/.

\begin{example}
Nasal place assimilation in Latinate prefixes:

>>>>

Above, it was assumed that the velar nasal is derived from underlying /n/. 
This analysis is not purely allophonic, as nasal place assimilation alternations target the final consonant of the \emph{im-}/\emph{in-} prefix. In (\ref{npa}ab), the prefix-final consonant takes on the same major place of articulation as the following obstruent, and in (\ref{npa}c), where the root is vowel-initial, the prefix-final consonant is realized as a default /n/.

\begin{example}
\label{npa}
Nasal place assimilation in Latinate prefixes:
